Spitzer, Gaia, and the Potential of the Milky Way

### 2013-08-08 text/paper1/paper.tex

Near-future data from ESA's Gaia mission will provide precise, full phase-space information for hundreds of millions of stars out to heliocentric distances of $\sim$ 10 kpc. This ``horizon'' for full phase-space measurements is imposed by the Gaia parallax errors degrading to worse than 10\%, and could be significantly extended by an accurate distance indicator. Recent work has demonstrated how Spitzer observations of RR Lyrae stars can be used to make distance estimates accurate to 2\%, effectively extending the Gaia, precise-data horizon by a factor of ten in distance and a factor of 1000 in volume. This \emph{Letter} presents one approach to exploit data of such accuracy to measure the Galactic potential using small samples of stars associated with debris from satellite destruction. The method is tested with synthetic observations of 100 stars from the end point of a simulation of satellite destruction: the shape, orientation, and depth of the potential used in the simulation are recovered to within a few percent. The success of this simple test with such a small sample in a single debris stream suggests that constraints from multiple streams could be combined to examine the Galaxy's  dark matter halo in even more  detail --- a truly unique opportunity that is enabled by the combination of Spitzer and Gaia with our intimate perspective on our own Galaxy.

@@

The Milky Way is the best candidate for such a detailed study of a dark matter halo
since we can resolve large samples of stellar tracers.
To date, thousands of blue horizontal branch stars selected from the Sloan Digital Sky Survey (SDSS) have been used to probe the Milky Way mass out to tens of
kpc \citep[SDSS, see][]{deason12a,kafle12}, and estimates with combined tracers extend to 150kpc \citep{deason12b}.

@@

A complimentary approach to measuring the mass distribution is to instead take full advantage of the {\it non}-random nature of the Galaxy's stellar distribution and utilize the knowledge that stars in streams were once all part of the same object.
Such approaches can require orders of magnitude fewer tracers than a randomly sampled population to achieve comparable accuracy.
One method is to simply fit orbits to observations of streams 
\citep[e.g.,][]{koposov10}.
However, the assumption that debris traces a single orbit is actually incorrect \citep[see][]{johnston98,helmi99}, with
systematic changes in orbital properties along debris streams that can lead to systematic biases in our measurements of the Galactic potential \citep{eyre09a,varghese11}.
In particular, \citet{sanders13a} recently demonstrated that this bias is equally problematic for the very thinnest, coldest streams, whose observed properties may be indistinguishable from those of the parent orbit \citep[e.g. such as the globular cluster, GD1 --- see][]{koposov10}, as for the much more extended and hotter streams \citep[e.g. such as debris from the Sagittarius dwarf galaxy --- see][]{majewski03} where offsets from a single orbit are clearly apparent.

@@

Other approaches have been proposed that avoid the cost of full N-body simulations yet recognize the limitations of single orbit integration.
For example, using the results of N-body simulations as a guide, simple corrections around a single orbit can be adopted to make model predictions that closely
mimic expected debris behavior \citep[e.g.][]{johnston99b,varghese11}.

The promise of near-future data sets including full phase-space information has also inspired other approaches. \citet{binney08} and \citet{penarrubia12} demonstrate conceptually that the distribution of energy and entropy in debris, respectively, will be minimized only for a correct
assumption of the form of the Galactic potential.
\citet{sanders13b} examine the distribution of debris in action-angle co-ordinates and show that stars stripped from the same disrupted object must lie along a single line in angle-frequency space, providing a constraint that can be used as a potential measure.

In this paper we re-examine and update a complimentary approach to using tidal debris as a potential measure \citep[originally proposed by][]{johnston99a}  in the context of current and near-future observational capabilities, and apply it to a simulation of the Sgr debris system.
In Section \ref{sec:context} we outline the observational prospects and Sgr properties that motivated this re-examination.
In Section \ref{sec:method} we present the updated potential measure and test it with synthetic observations of simulated Sgr debris.
In Section \ref{sec:discussion} we highlight the advantages and shortcomings of this method.
We conclude in Section \ref{sec:conclusion}.

@@

At longer wavelengths, RR Lyrae promise tighter constraints on
distances.  
\citet{madore12} have recently shown, 
using five stars with
trigonometric parallaxes measured by Hubble \citep{benedict11},
that the dispersion in the mid-IR Period-Luminosity relation 
\citep[first mapped by][]{longmore86}
at
wavelengths measurable by NASA's {\it Spitzer} mission is $\sim$0.03~mag.
This implies that it is
possible to use {\it Spitzer} to determine distances that are good to better
than $2\%$ for individual RR Lyrae stars out to $\sim$60~kpc (the limit of
Spitzer's ability to detect and measure RR Lyrae variables for modest
integration times). 
For comparison, distance measurements of Blue Horizontal Branch
stars typically
achieve $\sim$10-15\% uncertainties  \citep[if appropriate color measurements are available, e.g.,][]{deason12b}.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.45\textwidth]{/Users/adrian/projects/streams/plots/paper1/gaia.pdf}
\caption{Expected \emph{Gaia} distance and tangential velocity errors as a function of heliocentric distance for RR Lyrae stars. Errors are a function of color and magnitude of the source, and hence the metallicity. Parallax distance errors from \emph{Gaia} are larger than the line-of-sight size of both Sgr and Orphan (Orp), but photometric distance errors are comparable to the the Sgr scale (assuming 10\% errors, dotted line). Bottom panel shows that the \emph{Gaia} tangential velocity errors are smaller than the internal velocity dispersion of both Sgr and Orp.\footnote{\url{http://www.rssd.esa.int/index.php?page=Science_Performance&project=GAIA}} }\label{fig:gaia_errors}
\end{center}
\end{figure}

\subsection{Gaia and the age of astrometry}
\label{sec:gaia}
The Gaia satellite \citep{gaia01} is
an astrometric mission which aims to measure the positions of billions
of stars with 10-100~$\mu$as accuracies. Combined with expected 
proper motion accuracies, this will enable full six-dimensional phase-space 
maps of the Galaxy with $<$10\% distance errors to heliocentric distances of
up to $\sim$6~kpc for RR Lyrae stars.

Figure~\ref{fig:gaia_errors} shows the Gaia end-of-mission distance
and tangential velocity error estimates for RR Lyrae. The velocity
uncertainty depends on the magnitude (and color) of the source, which depends in turn on
the metallicity: each line in Figure~\ref{fig:gaia_errors} is computed
by Monte Carlo sampling from the empirical metallicity distribution of
the Galactic halo from \cite{ivezic08} (centered around [Fe/H] $\sim$
-1.5). Within 2~kpc, Gaia will measure distances to these stars with
better than 2\% accuracy --- RR Lyrae in this volume can be used to
test and calibrate the Spitzer PL relation described above. Beyond the
2~kpc threshold, the mid-IR PL relation for RR Lyrae will provide better 
distance measurements.

@@

\subsection{The Sagittarius debris system}
\label{sec:sgr}
The Sagittarius Dwarf Galaxy (hereafter Sgr) was discovered
serendipitously during a radial velocity survey of the Galactic bulge
\citep{ibata94}. 
Signatures of extensive stellar
streams associated with Sgr have since  been
mapped across the sky in carbon stars \citep{totten98}, M giants
selected from the Two Micron All Sky Survey \citep{majewski03}, main
sequence turnoff stars from SDSS
\citep{belokurov06}, and RR Lyrae in the Catalina Sky Survey
\citep{drake13}.

@@

LM10 demonstrate how a comparison of such simulations with the
observed debris enables a detailed reconstruction of Sgr's history:
its mass and stellar populations, orbit, rate of destruction, and even
original distribution in populations. 
Moreover, this comparison enabled the first 
assessment of the three-dimensional mass distribution of a dark matter
halo through the constraints on the potential parameters $v_{\rm halo}$, $q_1$, $q_z$, and $\phi$.

Combined Spitzer and Gaia measurements of distances and proper motions
of RR Lyraes in the Sgr debris will open up new avenues of
interpretation. A 2\% distance error at $\sim$30~kpc
($\epsilon_{d}=0.6~{\rm kpc}$) is smaller than the predicted thickness
of the stream, ranging from $\sim$2-$10$~kpc. When combined with Gaia proper motion errors,
tangential velocity measurement accuracies ($\epsilon_{v}\sim2$~km/s
at 30~kpc; see Figure~\ref{fig:gaia_errors}) are less than the intrinsic dispersion for much of the stream
\citep[$\sigma_v\sim10$~km/s;][]{majewski04} and the measured phase-space position of each individual RR
Lyrae becomes a powerful probe of the potential. The next section
outlines a new method to take advantage of this information.

\section{Description and test of our algorithm}
\label{sec:method}
With access to 6D information for stars in a tidal
stream, it is possible to use the individual stars as potential
measures by exploiting the fact that the stars must have come from the
same progenitor: if the orbits of the stars and progenitor are integrated 
\emph{backwards} in a a potential that accurately models the Milky Way
(imagine watching satellite destruction in ``rewind"), the stars
should recombine with the progenitor. If the potential is incorrect,
the orbits of the stars will diverge from that of the progenitor and
thus will not be recaptured by the satellite system (Figure~\ref{fig:ps_distance}).

@@

@@

%\begin{figure}[h]
%\begin{center}
%\includegraphics[width=0.45\textwidth]{/Users/adrian/projects/streams/plots/paper1/bootstrap.pdf}
%\caption{ Blue points show the ``best-fit'' parameters resulting from each bootstrap resample of 100 stars from the Sgr stream particle density shown in Figure~\ref{fig:lm10}. Green (vertical and horizontal) lines show the true values of the parameters. Grey ellipses show one- and two-sigma margins, assuming the points are normally distributed. }\label{fig:bootstrap}
%\end{center}
%\end{figure}

\begin{figure*}[!t]
\centering\includegraphics[width=0.9\textwidth,trim=0 0 0 0, clip]{/Users/adrian/projects/streams/plots/paper1/bootstrap.pdf}
\caption{ Blue points show the ``best-fit'' parameters resulting from each resample of 100 stars from the Sgr stream particle density shown in Figure~\ref{fig:lm10}. Green (vertical and horizontal) lines show the true values of the parameters. Grey ellipses show one- and two-sigma margins, assuming the points are normally distributed. }\label{fig:bootstrap}
\end{figure*}

>>>>

Near-future data from ESA's Gaia mission will provide precise, full phase-space information for hundreds of millions of stars out to heliocentric distances of $\sim$10 kpc. This ``horizon'' for full phase-space measurements is imposed by the Gaia parallax errors degrading to worse than 10\%, and could be significantly extended by an accurate distance indicator. Recent work has demonstrated how Spitzer observations of RR Lyrae stars can be used to make distance estimates accurate to 2\%, effectively extending the Gaia, precise-data horizon by a factor of ten in distance and a factor of 1000 in volume. This \emph{Letter} presents one approach to exploit data of such accuracy to measure the Galactic potential using small samples of stars associated with debris from satellite destruction. The method is tested with synthetic observations of 100 stars from the end point of a simulation of satellite destruction: the shape, orientation, and depth of the potential used in the simulation are recovered to within a few percent. The success of this simple test with such a small sample in a single debris stream suggests that constraints from multiple streams could be combined to examine the Galaxy's  dark matter halo in even more  detail --- a truly unique opportunity that is enabled by the combination of Spitzer and Gaia with our intimate perspective on our own Galaxy.

@@

The Milky Way is the best candidate for such a detailed study of a dark matter halo
since we can resolve large samples of stellar tracers.
Thousands of blue horizontal branch stars selected from the Sloan Digital Sky Survey (SDSS) have been used to probe the Milky Way mass out to tens of
kpc \citep[SDSS, see][]{deason12a,kafle12}, and estimates with combined tracers extend to 150kpc \citep{deason12b}.

@@

A complimentary approach to measuring the mass distribution is to instead take advantage of the {\it non}-random nature of the Galaxy's stellar distribution and utilize the knowledge that stars in streams were once all part of the same object.
Such approaches can require orders of magnitude fewer tracers than a randomly sampled population to achieve comparable accuracy.
One method is to simply fit orbits to observations of streams 
\citep[e.g.,][]{koposov10}.
However, the assumption that debris traces a single orbit is actually incorrect \citep[see][]{johnston98,helmi99}.
Changes in orbital properties along debris streams that can lead to systematic biases in measurements of the Galactic potential \citep{eyre09a,varghese11}.
In particular, \citet{sanders13a} recently demonstrated that this bias is equally problematic for the very thinnest, coldest streams, whose observed properties may be indistinguishable from those of the parent orbit \citep[e.g. such as the globular cluster, GD1 --- see][]{koposov10}, as for the much more extended and hotter streams \citep[e.g. such as debris from the Sagittarius dwarf galaxy --- see][]{majewski03} where offsets from a single orbit are clearly apparent.

@@

The promise of near-future data sets including full phase-space information has also inspired other approaches. \citet{binney08} and \citet{penarrubia12} demonstrate that the distribution of energy and entropy in debris, respectively, will be minimized only for a correct
assumption of the form of the Galactic potential.
\citet{sanders13b} examine the distribution of debris in action-angle co-ordinates and show that stars stripped from the same disrupted object must lie along a single line in angle-frequency space, providing a constraint that can be used as a potential measure.

In this \emph{Letter} we re-examine and update a complimentary approach to using tidal debris as a potential measure \citep[originally proposed by][]{johnston99a}  in the context of current and near-future observational capabilities, and apply it to a simulation of the Sgr debris system.
In Section \ref{sec:context} we outline the observational prospects and Sgr properties that motivated this re-examination.
In Section \ref{sec:method} we present the updated potential measure and test it with synthetic observations of simulated Sgr debris.
In Section \ref{sec:discussion} we highlight the advantages and shortcomings of this method.
We conclude in Section \ref{sec:conclusion}.

@@

At longer wavelengths, RR Lyrae promise tighter constraints on
distances.  
\citet{madore12} have recently shown, 
using five stars with
trigonometric parallaxes measured by Hubble \citep{benedict11},
that the dispersion in the mid-IR Period-Luminosity (PL) relation 
\citep[first mapped by][]{longmore86}
at
wavelengths measurable by NASA's {\it Spitzer} mission is $\sim$0.03~mag.
This implies that it is
possible to use {\it Spitzer} to determine distances that are good to $2\%$ for 
individual RR Lyrae stars out to $\sim$60~kpc ({\it Spitzer}'s limit for detecting and measuring RR Lyrae).
For comparison, distance measurements of Blue Horizontal Branch
stars typically
achieve $\sim$10-15\% uncertainties  \citep[if appropriate color measurements are available, e.g.,][]{deason12b}.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.45\textwidth]{/Users/adrian/projects/streams/plots/paper1/gaia.pdf}

\caption{Expected \emph{Gaia} distance and tangential velocity errors as a function of heliocentric distance for RR Lyrae stars. Errors are a function of color and magnitude of the source, and hence the metallicity: each line is computed
by Monte Carlo sampling from the empirical metallicity distribution of
the Galactic halo from \cite{ivezic08}. Parallax distance errors from \emph{Gaia} are larger than the line-of-sight size of both Sgr and Orphan (Orp), but photometric distance errors are comparable to the the Sgr scale (assuming 10\% errors, dotted line). Bottom panel shows that the \emph{Gaia} tangential velocity errors are smaller than the internal velocity dispersion of both Sgr and Orp.\footnote{\url{http://www.rssd.esa.int/index.php?page=Science_Performance&project=GAIA}} }\label{fig:gaia_errors}
\end{center}
\end{figure}

\subsection{Gaia and the age of astrometry}
\label{sec:gaia}
The Gaia satellite \citep{gaia01} is
an astrometric mission which aims to measure the positions of billions
of stars with 10-100~$\mu$as accuracies. Combined with expected 
proper motion accuracies, this will enable full six-dimensional phase-space 
maps of the Galaxy with $<$10\% distance errors for heliocentric distances of
up to $\sim$6~kpc for RR Lyrae stars.

Figure~\ref{fig:gaia_errors} shows the Gaia end-of-mission distance
and tangential velocity error estimates for RR Lyrae. Within 2~kpc, Gaia will measure distances to these stars with
better than 2\% accuracy --- RR Lyrae in this volume can be used to
test and calibrate the Spitzer PL relation described above. Beyond the
2~kpc threshold, the mid-IR PL relation for RR Lyrae will provide better 
distance measurements.

@@

\subsection{The Sagittarius debris system}
\label{sec:sgr}
The Sagittarius Dwarf Galaxy (hereafter Sgr) was discovered
serendipitously during a radial velocity survey of the Galactic bulge
\citep{ibata94}. 
Signatures of extensive stellar
streams associated with Sgr have since  been
mapped across the sky in carbon stars \citep{totten98}, M giants
selected from 2MASS \citep{majewski03}, main
sequence turnoff stars from SDSS
\citep{belokurov06}, and RR Lyrae in the Catalina Sky Survey
\citep{drake13}.

@@

LM10 demonstrate how a comparison of such simulations with the
observed debris enables a detailed reconstruction of Sgr's history:
its mass and stellar populations, orbit, rate of destruction, and even
original distribution in populations. 
This comparison enabled the first 
assessment of the three-dimensional mass distribution of a dark matter
halo through the constraints on the potential parameters $v_{\rm halo}$, $q_1$, $q_z$, and $\phi$.

Combined {\it Spitzer} and {\it Gaia} measurements of distances and proper motions
of RR Lyrae in the Sgr debris will open up new avenues of
interpretation. A 2\% distance error at $\sim$30~kpc
($\epsilon_{d}=0.6~{\rm kpc}$) is smaller than the predicted thickness
of the stream ($\sim$2-$10$~kpc). When combined with Gaia proper motion errors,
tangential velocity measurement accuracies ($\epsilon_{v}\sim2$~km/s
at 30~kpc; see Figure~\ref{fig:gaia_errors}) are less than the intrinsic dispersion for much of the stream
\citep[$\sigma_v\sim10$~km/s;][]{majewski04}. The next section
outlines a new method to take advantage of this information.

\section{Description and test of our algorithm}
\label{sec:method}
With access to 6D information for stars in a tidal
stream, each star becomes a powerful potential
measure by exploiting the fact that the stars must have come from the
same progenitor: if the orbits of the stars and progenitor are integrated 
\emph{backwards} in a a potential that accurately models the Milky Way
(imagine watching satellite destruction in ``rewind"), the stars
should recombine with the progenitor. If the potential is incorrect,
the orbits of the stars will diverge from that of the progenitor and
thus will not be recaptured by the satellite system (Figure~\ref{fig:ps_distance}).

@@

\begin{figure*}[ht!]
\centering\includegraphics[width=0.9\textwidth,trim=0 0 0 0, clip]{/Users/adrian/projects/streams/plots/paper1/bootstrap.pdf}
\caption{ Blue points show the ``best-fit'' parameters resulting from each resample of 100 stars from the Sgr stream particle density shown in Figure~\ref{fig:lm10}. Green (vertical and horizontal) lines show the true values of the parameters. Grey ellipses show one- and two-sigma margins, assuming the points are normally distributed. }\label{fig:bootstrap}
\end{figure*}

@@
